Cover letter for the original IJRR submission

### 2017-05-03 body/abstract.tex
Tentative complete paper.

% !TEX root =  ../main.tex
We present a contact planner for complex legged locomotion tasks: standing up, climbing stairs using a handrail, crossing rubble and getting out of a car. The need for such a planner was shown at the Darpa Robotics Challenge, where such behaviors
could not be demonstrated (except for egress).
%
%This planning problem is theoretically known to be a specific instance of a larger class of complex planning problems (``manipulation'' or ``rearrangement'' planning).
Current planners suffer from prohibitive algorithmic complexity, because they deploy a tree of robot configurations projected in contact with the environment.
Contact configurations are costly to compute (because sampled and projected in high dimension) and have no equivalence property (i.e. two neighbor configurations cannot be directly connected in the planning graph).
%
We tackle this issue by introducing a reduction property: the reachability condition. 
This condition defines a geometric approximation of the contact manifold, which is of low dimension, has Cartesian topology and can be efficiently sampled and explored.
%
The hard contact planning problem can then be separated into two sub-problems: 
first, we plan a path for the root without considering the whole-body configuration, using a sampling-based algorithm; 
second, we generate a discrete sequence of whole-body configurations in static equilibrium along this path, using a deterministic contact-selection algorithm. 
%
%We tackle this issue by introducing the reachability condition. This condition allows us to define
%a geometric approximation of the contact manifold, which is of low dimension and can be sampled efficiently.
%Informally, the condition verifies that the root configuration of a robot is close, but not too close to obstacles: close to allow contact creation, not too close to avoid collision. Thanks to this condition we decompose the hard contact planning problem into two
%sub-problems: first, planning a guide path for the root without considering the whole-body configuration, using a sampling-based algorithm; then, generating a discrete sequence of whole-body configurations in static equilibrium along this path, using a deterministic contact-selection algorithm. 
%
The reduction breaks the algorithm complexity encountered in other contact planners, logically leading to the open-source implementation of the first interactive multi-contact planner.
%
While no contact planner has yet been proposed with theoretical completeness, we empirically show by a large number of trials ($>$10,000) that it is not an issue for our method:
in a few seconds, with high success rates, we generate complex contact plans for various scenarios and robots: HRP-2, HyQ and a dexterous hand. 
These plans are then validated either in dynamic simulations and on the real HRP-2 robot.
%, thus demonstrating the first interactive multi-contact planner.
%Our approach results from the pragmatic choice of favoring efficiency over completeness, which we justify empirically

>>>>

% !TEX root =  ../main.tex
We present a contact planner for complex legged locomotion tasks: standing up, climbing stairs using a handrail, crossing rubble, getting out of a car. The need for such a planner was shown at the Darpa Robotics Challenge, where such behaviors
could not be demonstrated (except for egress).
%
%This planning problem is theoretically known to be a specific instance of a larger class of complex planning problems (``manipulation'' or ``rearrangement'' planning).
Current planners suffer from prohibitive algorithmic complexity, because they deploy a tree of robot configurations projected in contact with the environment.
%Contact configurations are costly to compute (because sampled and projected in high dimension) and have no equivalence property (i.e. two neighbor configurations cannot be directly connected in the planning graph).
%
We tackle this issue by introducing a reduction property: the reachability condition. 
This condition defines a geometric approximation of the contact manifold, which is of low dimension, has Cartesian topology and can be efficiently sampled and explored.
%
The hard contact planning problem can then be separated into two sub-problems: 
first, we plan a path for the root without considering the whole-body configuration, using a sampling-based algorithm; 
second, we generate a discrete sequence of whole-body configurations in static equilibrium along this path, using a deterministic contact-selection algorithm. 
%
%We tackle this issue by introducing the reachability condition. This condition allows us to define
%a geometric approximation of the contact manifold, which is of low dimension and can be sampled efficiently.
%Informally, the condition verifies that the root configuration of a robot is close, but not too close to obstacles: close to allow contact creation, not too close to avoid collision. Thanks to this condition we decompose the hard contact planning problem into two
%sub-problems: first, planning a guide path for the root without considering the whole-body configuration, using a sampling-based algorithm; then, generating a discrete sequence of whole-body configurations in static equilibrium along this path, using a deterministic contact-selection algorithm. 
%
The reduction breaks the algorithm complexity encountered in previous works, logically leading to the  first interactive implementation of a contact planner (open source).
%
While no contact planner has yet been proposed with theoretical completeness, we empirically show by a large number of trials ($>$10,000) that it is not an issue for our method:
in a few seconds, with high success rates, we generate complex contact plans for various scenarios and robots: HRP-2, HyQ and a dexterous hand. 
These plans are validated either in dynamic simulations and on the real HRP-2 robot.
%, thus demonstrating the first interactive multi-contact planner.
%Our approach results from the pragmatic choice of favoring efficiency over completeness, which we justify empirically

### 2017-05-03 body/conclusion.tex
Tentative complete paper.

As demonstrated in the results section, the main interest of our method is time performance. Time performance is critical for 
online applications with the robot, a goal not satisfied by any of the previous contributions. Our method addresses highly constrained environments while improving the search time by orders of magnitude. This high performance is reached at the cost of some approximations that we discuss here. 
%~ In previous approaches, we identified 2 sources of high computational cost.

The first approximation is the verification of \contactreachability\ ($\mathbf{q}^0 \in C_{Contact}^0$).  Our \textit{reachability condition} ($\mathbf{q}^0 \in C_{Reach}^0$) is computationally efficient and provides an accurate approximation of $C_{Contact}^0$ (Section~\ref{sec:scaling}). This is demonstrated by the second column of Table~\ref{tab:requestpercent}, and illustrated by Figure~\ref{fig:dedefeas}. Indeed, in the large majority of cases, (84\% in the worst car egress case), we are able to find a contact configuration for any configuration in $C_{Reach}^0$.

@@

These results clearly justify our pragmatic approach.

\subsection{Addressing $\mathcal{P}_3$}
This paper focuses on contact planning of multi-contact locomotion.It does not address directly the continuous motion generation,
. Problem $\mathcal{P}_3$ remains an active research topic~\cite{Carpentier2016, herzog2015trajectory},
for which no off-the shelf solution exists thus far. We thus implemented a dedicated solution, described in Appendix~\ref{app:optim}.
The companion video shows the results that we obtained in simulation and on the real HRP-2 robot.
In these scenarios the robot is able to execute our plans, while respecting the constraints inherent to the problem: joint kinematics and torque limits and
collision avoidance.
%~ \adnote{Probably you should add in appendix an explanation of how you solve $P_3$ and refer to it here.}

@@

In this paper we consider the multi-contact planning problem, formulated as two sub-problems that we address sequentially.
The first problem $\mathcal{P}_1$ consists in computing an \gls{equilibrium feasible} guide path for the root of the robot;
 the second problem $\mathcal{P}_2$ is the computation of a discrete sequence of whole-body configurations along the root path.

Our contribution to \Pa is the introduction of a low-dimensional space $C_{reach}^0$, an approximation of the space of \gls{equilibrium feasible} root configurations.
$C_{reach}^0$ can be efficiently sampled and has a low-dimension. For these reasons we are able to solve \Pa much faster than previous approaches.

%~ Our contribution to \Pa is a generic characterization of the properties that the guide path must satisfy, in particular to enforce the completeness of the acyclic contact planner. 
Our contribution to \Pb is a fast contact generation scheme that can optimize user-defined criteria.

Our results demonstrate that our method allows a pragmatic compromise between three 
criteria that are hard to reconcile: generality, performance, and quality of the solution, making it the first acyclic contact
planner compatible with \gls{interactive} applications.
%
\textbf{Regarding generality}, the \textit{reachability condition}, coupled with an approach based on limb decomposition, 
allows the method to address automatically arbitrary legged robots.
%
\textbf{Regarding performance}, our framework is efficient in addressing both \Pa and $\mathcal{P}_2$. This results in \gls{interactive} computation times.
%
\textbf{Regarding the quality of the paths}, we are able to compute
\gls{equilibrium feasible} paths in all the presented scenarios, with high success rates.
As for \cite{Bouyarmane2009}, failures can still occur, due to the approximate condition used to compute the guide path.
The low computational burden of our framework however allows for fast re-planning in case of failure.
Furthermore, because of this approximation, the guide search is not complete. The choice is deliberate, because we believe
that it is necessary to trade completeness for efficiency at all stages of the planner.
However, one direction for future work is to focus on a more accurate formulation of $C_{reach}^0$ to improve
the approximation.

>>>>

As demonstrated in the results section, the main interest of our method is reduced algorithmic complexity, leading to interactive implementation. This property is critical for 
online applications with the robot and was not reached by any of the previous contributions. Our method addresses highly constrained environments while improving the search time by orders of magnitude. This high performance is reached at the cost of some approximations that we discuss here. 
%~ In previous approaches, we identified 2 sources of high computational cost.

The first approximation occurs when verifying the \contactreachability\ ($\mathbf{q}^0 \in C_{Contact}^0$).  Our \textit{reachability condition} ($\mathbf{q}^0 \in C_{Reach}^0$) is computationally efficient and provides an accurate approximation of $C_{Contact}^0$ (Section~\ref{sec:scaling}). This is demonstrated by the second column of Table~\ref{tab:requestpercent}, and illustrated by Figure~\ref{fig:dedefeas}. Indeed, in the large majority of cases, (84\% in the worst car egress case), we are able to find a contact configuration for any configuration in $C_{Reach}^0$.

@@

These results clearly justify our pragmatic approach.
Moreover, refining the reachibility condition to makes it more accurate while keeping the separated approach seems a fruitful research direction.

\subsection{Addressing $\mathcal{P}_3$}
This paper focuses on contact planning of multi-contact locomotion. Following the spirit of previous works~\cite{Bretl:2006:MPM:1124573.1124585,DBLP:conf/iser/EscandeKMG08,7140082},
it does not address directly the continuous motion generation. Problem $\mathcal{P}_3$ remains an active research topic~\cite{Carpentier2016, herzog2015trajectory},
for which no off-the shelf solution exists thus far. We thus implemented a dedicated solution, described in Appendix~\ref{app:optim}.
The companion video shows the results that we obtained in simulation and on the real HRP-2 robot.
In these scenarios the robot is able to execute our plans, while respecting the constraints inherent to the problem: joint kinematics and torque limits and
collision avoidance.
%~ \adnote{Probably you should add in appendix an explanation of how you solve $P_3$ and refer to it here.}

@@

In this paper we consider the multi-contact planning problem, formulated as two sub-problems that we address sequentially.
The first problem $\mathcal{P}_1$ consists in computing an \gls{equilibrium feasible} guide path for the root of the robot;
 the second problem $\mathcal{P}_2$ is the computation of a discrete sequence of whole-body configurations along the root path.
%
Our contribution to \Pa is the introduction of a low-dimensional space $C_{reach}^0$, an approximation of the space of \gls{equilibrium feasible} root configurations.
$C_{reach}^0$ can be efficiently sampled and has a low-dimension. For these reasons we are able to solve \Pa much faster than previous approaches.
%
%~ Our contribution to \Pa is a generic characterization of the properties that the guide path must satisfy, in particular to enforce the completeness of the acyclic contact planner. 
Our contribution to \Pb is a fast contact generation scheme that can optimize user-defined criteria.

Our results demonstrate that our method allows a pragmatic trade-off between three 
criteria that are hard to reconcile: generality, performance, and quality of the solution, making it the first acyclic contact
planner compatible with \gls{interactive} applications.
%
\textbf{Regarding generality}, the \textit{reachability condition}, coupled with an approach based on limb decomposition, 
allows the method to address automatically arbitrary legged robots.
%
\textbf{Regarding performance}, our framework is efficient in addressing both \Pa and $\mathcal{P}_2$. This results in \gls{interactive} computation times.
%
\textbf{Regarding the quality of the paths}, we are able to compute
\gls{equilibrium feasible} paths in all the presented scenarios, with high success rates.
As for \cite{Bouyarmane2009}, failures can still occur, due to the approximate condition used to compute the guide path.
The low computational burden of our framework however allows for fast re-planning in case of failure.
Furthermore, because of this approximation, the guide search is not complete. The choice is deliberate, because we believe
that it is necessary to trade completeness for efficiency at all stages of the planner.
However, one direction for future work is to focus on a more accurate formulation of $C_{reach}^0$ to improve
the approximation.

### 2017-05-03 body/contact.tex
Tentative complete paper.

First, the root path $\mathbf{q}^0(t)$ is discretized into a sequence of $j$ key configurations:  
\begin{equation*}
	\mathbf{Q}^0 = [\mathbf{q}^0_{0}; \mathbf{q}^0_{i}; ..., \mathbf{q}^0_{j-1}]
\end{equation*} 
where $\mathbf{q}^0_{0}$ and $\mathbf{q}^0_{j-1}$ are the start and goal configurations. %To ensure continuity in the contact transition phases,
Each root configuration of $\mathbf{Q}^0$ is then extended into a whole-body configuration such that:
\begin{itemize} 
\item At most one contact is not maintained (\textit{broken}) between two consecutive configurations.
\item At most one contact is added between two consecutive configurations.
\item Each configuration is in static equilibrium.
\item Each configuration is collision-free.
\end{itemize}

@@

%~ Figure~\ref{fig:break_contact} illustrates the contact-persistence strategy.
If kinematically possible, a limb in contact at step $i-1$ remains in contact at step $i$ (Figure~\ref{fig:break_contact}). 
%~ The contact is broken if an inverse-kinematics solver fails to find a collision-free limb configuration that satisfies joint limits. 
%~ The solver is directly provided by the HPP software.
Otherwise the contact is broken and a collision-free configuration is assigned to the limb.
If two or more contacts can't be maintained between two consecutive configurations, one or more intermediate configurations are added, to ensure
that at most one contact is broken between two sequential configurations.
%~ For these steps the root configuration is the same as for the previous step, with the difference that
%~ one faulty contact is repositioned, in the hope that it will not be broken at the next step.

@@

To generate contacts, we proceed as follows.
The contact generator treats each limb $R^k$ independently.
Given a configuration of the root and the other limbs, it computes a configuration $\mathbf{q}^k$ such that $R^k$ is in contact and the robot $R$ is in static equilibrium:
\begin{equation}
\label{eq:contact_gen}
	\mathbf{q}^{\overline{k}}  \longrightarrow \mathbf{q}^k, (\mathbf{q}^{k} \oplus \mathbf{q}^{\overline{k}}) \in  C_{Equil} \textrm{ \textbf{and}}\ \mathbf{q}^k \in  C_{Contact}^k 
\end{equation}

Contact generation is typically adressed by randomly sampling a limb configuration, before projecting the effector onto the closest surface with an inverse kinematics solver.
This process is repeated until either a solution for Eq.~(\ref{eq:contact_gen}) is found, or the generator fails, according to a user-defined maximum number of trials.
We favor a modified implementation of this naive approach, more computationally efficient, introduced in our previous work~\citep{Tonneau2014}.

%~ Given a configuration $\mathbf{q}^{\overline{k}}$ of the root and all the limbs but $R^k$, we look for a limb configuration $\mathbf{q}^k$ such that
%~ $R^k$ is in contact, and not colliding (neither with parts of the robot nor with the environment).
%~ While exhibiting analytically a $\mathbf{q}^{k}$ does not seem tractable, we can iteratively try to generate one as follows:
%~ \begin{enumerate}
%~ \item Generate randomly a collision-free limb configuration;
%~ \item Project the end-effector onto the closest surface with inverse kinematics;
%~ \item If a valid solution is found, stop. Otherwise repeat from step 1.
%~ \end{enumerate}

%~ We favor a modified implementation of this naive approach, more computationally efficient, introduced in our previous work~\citep{Tonneau2014}.

We define $C_{Contact}^{\epsilon} \supset C_{Contact}$ as the set of configurations such that the minimum distance 
between an effector and an obstacle is less than $\epsilon \in \mathbb{R}$.
We then apply the following steps:
\begin{enumerate}
\item Generate off-line $N$ valid sample limb configurations $\mathbf{q}^k_i,  0 \leq i < N$ (We choose $N=10^3$);
\item Using the end-effector positions $\mathbf{p}(\mathbf{q}^k_i)$ as indices, store each sample in an octree data structure;
\item At runtime, when contact creation is required, intersect the octree and the environment to retrieve the list of samples $S \subset C_{Contact}^{\epsilon}$ close to contact (Figure~\ref{fig:contact_gen} (b) and (c));
\item Use a user-defined heuristic $h$ to sort $S$;
\item If $S$ is empty, stop (failure). Else select the first configuration of $S$. Project it onto contact using inverse kinematics. (Figure~\ref{fig:contact_gen} (d) and (e));
\item If Eq.~\ref{eq:contact_gen} is verified, stop (success). Otherwise remove the element from $S$ and go to step 5.
\end{enumerate}

%~ This approach has two main advantages.
%~ First, this allows us to select a large number of candidates in a single proximity request.
%~ Having several candidates is interesting, because it allows to compare them using a user-selected heuristic $h$, thus obtaining
%~ a locally-optimal candidate.
%~ Furthermore, the fact that the candidates are already close to contact increases the odds that the inverse kinematics will converge to a valid solution in a small number of iterations.
%~ Regarding convergence, it is immediate to verify that as $N$ grows, the probability of finding a solution if it exists converges to 1.
%~ $N$ is a parameter that allows us to specify the trade-off between exhaustiveness and efficiency.
The reader is invited to refer to our previous work for an extensive discussion on the benefits of this approach, and the optimal choice 
of the parameter $N$~\citep{Tonneau2014}. Our criterion to assert efficiently the static equilibrium of the robot, as well as the heuristics $h$ are detailed in Appendix~\ref{sec:heuristics}.

>>>>

An initial discretization of the root path $\mathbf{q}^0(t)$ into a sequence of $j$ key configurations is first chosen:  
\begin{equation*}
	\mathbf{Q}^0 = [\mathbf{q}^0_{0}; ...; \mathbf{q}^0_{i}; ...; \mathbf{q}^0_{j-1}]
\end{equation*} 
where $\mathbf{q}^0_{0}$ and $\mathbf{q}^0_{j-1}$ are the start and goal configurations. %To ensure continuity in the contact transition phases,
Each root configuration $\mathbf{q}^0_i$ of $\mathbf{Q}^0$ is then extended into a whole-body configuration, while possibly additional discrete steps are added, such that:
\begin{itemize} 
\item At most one contact is not maintained (\textit{broken}) between two consecutive configurations.
\item At most one contact is added between two consecutive configurations.
\item Each configuration is in static equilibrium.
\item Each configuration is collision-free.
\end{itemize}

@@

%~ Figure~\ref{fig:break_contact} illustrates the contact-persistence strategy.
If kinematically possible, a limb in contact at step $i-1$ remains in contact at step $i$ (Figure~\ref{fig:break_contact}). 
%~ The contact is broken if an inverse-kinematics solver fails to find a collision-free limb configuration that satisfies joint limits. 
%~ The solver is directly provided by the HPP software.
Otherwise the contact is broken and a collision-free configuration is assigned to the limb.
If two or more contacts cannot be maintained between two consecutive configurations, one or more intermediate configurations are added, to ensure
that at most one contact is broken between two sequential configurations.
%~ For these steps the root configuration is the same as for the previous step, with the difference that
%~ one faulty contact is repositioned, in the hope that it will not be broken at the next step.

@@

Given a configuration of the root and the list of effectors that should be in contact, the contact generator computes the configuration of the limbs such that contacts are properly satisfied and the robot is in static equilibrium.
In previous works~\cite{DBLP:conf/iser/EscandeKMG08,Bouyarmane2009}, the generation of contact is typically implemented by randomly sampling limbs configurations and projecting the whole robot configuration onto the closest surfaces with an inverse kinematics solver.
In case of failure of the projection, the process would randomly iterate.

We propose two modifications of this general algorithm principle.
First our contact generator handle each limb $R^k$ independently.
By handling each limb separatly, we reduce the complexity of the generation of contact configurations.
This is made possible thanks to the reachibility condition in $\mathcal{P}_1$ that produces a root path that we can afford not to modify in $\mathcal{P}_2$, and because we allow both a contact break and a contact creation between two consecutive configurations of the contact sequence.
%
%% Given a configuration $\mathbf{q}^{\overline{k}}$ of the root and the other limbs, it computes a configuration $\mathbf{q}^k$ such that limb $R^k$ is in contact and the robot $R$ is in static equilibrium:
%% \begin{equation}
%% \label{eq:contact_gen}
%% 	\mathbf{q}^{\overline{k}}  \longrightarrow \mathbf{q}^k,  s.t.\ \mathbf{q}^k \in  C_{Contact}^k  \textrm{ \textbf{and}}\  (\mathbf{q}^{k} \oplus \mathbf{q}^{\overline{k}}) \in  C_{Equil}
%% \end{equation}
%% The limb configuration $\mathbf{q}^{\overline{k}}$ also defines the location of the contact points.
%
%% In previous work, the generation of contact configuration is typically achieved by randomly sampling a limb configuration, before projecting the effector onto the closest surface with an inverse kinematics solver, and iterating until
%
%% The contact generator is implemented by randomly sampling a limb configuration, before projecting the effector onto the closest surface with an inverse kinematics solver.
%% This process is repeated until either a solution for \eqref{eq:contact_gen} is found, or the generator fails, according to a user-defined maximum number of trials.
%% We favor a modified implementation of this naive approach, more computationally efficient, introduced in our previous work~\citep{Tonneau2014}.
%
%~ Given a configuration $\mathbf{q}^{\overline{k}}$ of the root and all the limbs but $R^k$, we look for a limb configuration $\mathbf{q}^k$ such that
%~ $R^k$ is in contact, and not colliding (neither with parts of the robot nor with the environment).
%~ While exhibiting analytically a $\mathbf{q}^{k}$ does not seem tractable, we can iteratively try to generate one as follows:
%~ \begin{enumerate}
%~ \item Generate randomly a collision-free limb configuration;
%~ \item Project the end-effector onto the closest surface with inverse kinematics;
%~ \item If a valid solution is found, stop. Otherwise repeat from step 1.
%~ \end{enumerate}
%
%~ We favor a modified implementation of this naive approach, more computationally efficient, introduced in our previous work~\citep{Tonneau2014}.
Second, we rely on off-line generation of configuration candidates.
We define $C_{Contact}^{\epsilon} \supset C_{Contact}$ as the set of configurations such that the minimum distance 
between an effector and an obstacle is less than $\epsilon \in \mathbb{R}$.
We then apply the following steps:
\begin{enumerate}
\item Generate off-line $N$ valid sample limb configurations $\mathbf{q}^k_i,  0 \leq i < N$ (we choose $N=10^3$);
\item Using the end-effector positions $\mathbf{p}(\mathbf{q}^k_i)$ as indices, store each sample in an octree data structure;
\item At runtime, when contact creation is required, intersect the octree and the environment to retrieve the list of samples $S \subset C_{Contact}^{\epsilon}$ close to contact (Figure~\ref{fig:contact_gen} (b) and (c));
\item Use a user-defined heuristic $h$ to sort $S$;
\item If $S$ is empty, stop (failure). Else select the first configuration of $S$. Project it onto contact using inverse kinematics (Figure~\ref{fig:contact_gen} (d) and (e));
\item If projections succeed and equilibrium is verified, stop (success). Otherwise remove the element from $S$ and go to step 5.
\end{enumerate}

%~ This approach has two main advantages.
%~ First, this allows us to select a large number of candidates in a single proximity request.
%~ Having several candidates is interesting, because it allows to compare them using a user-selected heuristic $h$, thus obtaining
%~ a locally-optimal candidate.
%~ Furthermore, the fact that the candidates are already close to contact increases the odds that the inverse kinematics will converge to a valid solution in a small number of iterations.
%~ Regarding convergence, it is immediate to verify that as $N$ grows, the probability of finding a solution if it exists converges to 1.
%~ $N$ is a parameter that allows us to specify the trade-off between exhaustiveness and efficiency.
The reader is refered to our previous work for an extensive discussion on the benefits of this approach, and the optimal choice 
of the parameter $N$~\citep{Tonneau2014}. Our criterion to assert efficiently the static equilibrium of the robot, as well as the heuristics $h$ are detailed in Appendix~\ref{sec:heuristics}.
